---
title: "Nelson-Aalen ties"
author: "Suzie Brown"
date: "13 November 2017"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)#

library(survival)
```

## Define functions
```{r}
makedata <- function(nobs=100, lambda=1, digits=2) { #sample the uncensored event times from exponential distn
  T <- rexp(nobs, lambda)
  T.rnd <- round(T, digits=digits)
  cbind(T, T.rnd)
}

censoring <- function(T, p=0.3){ # censor u.a.r.
  T.cens <- sample(length(T), floor(length(T)*p))
  delta <- rep(1, length(T))
  delta[T.cens] <- 0
  delta
}
```

## No ties
```{r}
y <- makedata()
delta <- censoring(y[,1])
ysurv <- Surv(y[,1], delta)
yfit <- summary(survfit(ysurv~1))

NAest <- cumsum(yfit$n.event/yfit$n.risk) # Nelson-Aalen estimator
NAvar <- cumsum(yfit$n.event/yfit$n.risk^2) # Nelson-Aalen variance estimator
```

### Plot results
```{r}
plot(yfit$time, NAest, col=4, type='S', xlab="time", ylab="cumulative hazard") # Nelson-Aalen estimator
points(yfit$time, -log(yfit$surv), type='S', col=2) # Kaplan-Meier (default in survfit)
points(yfit$time, yfit$time, type='l', col='grey', lty=2) # true hazard function (for lambda=1)
legend("topleft", c("Nelson-Aalen", "Kaplan-Meier", "true"), lty=c(1,1,2), col=c(4,2,'grey'))

## CI for survfit line
points(yfit$time, -log(yfit$surv + yfit$std.err), type='S', col=2, lty=3)
points(yfit$time, -log(yfit$surv - yfit$std.err), type='S', col=2, lty=3)

## CI for N-A line
points(yfit$time, NAest + NAvar^(1/2), type='S', col=4, lty=3)
points(yfit$time, NAest - NAvar^(1/2), type='S', col=4, lty=3)
```

## Now try with rounded values...
```{r}
length(unique(y[,2]))/length(y[,2]) # proportion of event/censoring times not tied
ysurv <- Surv(y[,2], delta)
yfit <- summary(survfit(ysurv~1))

NAest1 <- cumsum(yfit$n.event/yfit$n.risk) # Nelson-Aalen estimator with approach 1
NAvar1 <- cumsum((yfit$n.risk-yfit$n.event)*yfit$n.event/yfit$n.risk^3) # Nelson-Aalen variance estimator with approach 1

NAestim2 <- function(surv) {
  terms <- numeric(length(surv$time))
  for (i in 1:length(surv$time)) {
    if (surv$n.event[i] == 1) {terms[i] <- 1/surv$n.risk[i]}
    else {
      d <- surv$n.event[i]
      terms[i] <- sum(1/(surv$n.risk[i]-0:(d-1)))
    }
  }
  cumsum(terms)
}
NAest2 <- NAestim2(yfit) # Nelson-Aalen estimator with approach 2

NAvaria2 <- function(surv) {
  terms <- numeric(length(surv$time))
  for (i in 1:length(surv$time)) {
    d <- surv$n.event[i]
    terms[i] <- sum(1/(surv$n.risk[i]-0:(d-1))^2)
  }
  cumsum(terms)
}
NAvar2 <- NAvaria2(yfit)
```

### Plot results
```{r}
plot(yfit$time, NAest1, type='S', xlab="time", ylab="cumulative hazard", col=4) # Nelson-Aalen estimator approach 1
points(yfit$time, NAest2, type='S', xlab="time", ylab="cumulative hazard", col='purple') # Nelson-Aalen estimator approach 2
points(yfit$time, -log(yfit$surv), type='S', col=2) # Kaplan-Meier (default in survfit)
points(yfit$time, yfit$time, type='l', col='grey', lty=2) # true hazard function (for lambda=1)
legend("topleft", c("Nelson-Aalen 1", "Nelson-Aalen 2", "Kaplan-Meier", "true"), lty=c(1,1,1,2), col=c(4,'purple' ,2,'grey'))
 
## CI for survfit line
points(yfit$time, -log(yfit$surv + yfit$std.err), type='S', col=2, lty=3)
points(yfit$time, -log(yfit$surv - yfit$std.err), type='S', col=2, lty=3)

## CI for NA1 line
points(yfit$time, NAest1 + NAvar1^(1/2), type='S', lty=3, col=4)
points(yfit$time, NAest1 - NAvar1^(1/2), type='S', lty=3, col=4)

## CI for NA2 line
points(yfit$time, NAest1 + NAvar2^(1/2), type='S', lty=3, col='purple')
points(yfit$time, NAest1 - NAvar2^(1/2), type='S', lty=3, col='purple')
```


### Closer look at variance estimates
```{r}
plot(NAvar1, type='l', col=4, log='y')
lines(NAvar2, col='purple')
```

## Varying sample size
```{r}
Nobs <- 20
p.cens <- 0.1

y <- makedata(nobs=Nobs)
delta <- censoring(y[,1], p.cens)
ysurv <- Surv(y[,2], delta)
yfit <- summary(survfit(ysurv~1))

NAest1 <- cumsum(yfit$n.event/yfit$n.risk) # Nelson-Aalen estimator with approach 1
NAvar1 <- cumsum((yfit$n.risk-yfit$n.event)*yfit$n.event/yfit$n.risk^3) # Nelson-Aalen variance estimator with approach 1
NAest2 <- NAestim2(yfit) # Nelson-Aalen estimator with approach 2
NAvar2 <- NAvaria2(yfit)

plot(yfit$time, NAest1, type='S', xlab="time", ylab="cumulative hazard", col=4) # Nelson-Aalen estimator approach 1
points(yfit$time, NAest2, type='S', xlab="time", ylab="cumulative hazard", col='purple') # Nelson-Aalen estimator approach 2
points(yfit$time, yfit$time, type='l', col='grey', lty=2) # true hazard function (for lambda=1)
legend("topleft", c("Nelson-Aalen 1", "Nelson-Aalen 2", "true"), lty=c(1,1,2), col=c(4,'purple' ,'grey'))
## CI for NA1 line
points(yfit$time, NAest1 + NAvar1^(1/2), type='S', lty=3, col=4)
points(yfit$time, NAest1 - NAvar1^(1/2), type='S', lty=3, col=4)
## CI for NA2 line
points(yfit$time, NAest1 + NAvar2^(1/2), type='S', lty=3, col='purple')
points(yfit$time, NAest1 - NAvar2^(1/2), type='S', lty=3, col='purple')
```

### Comparative plots
This is not a good bit.
Changing `p.cens` doesn't affect the difference between estimators.

Difference of two N-A estimators:
```{r}
par(mfrow=c(1,3))
p.cens <- 0.3
y <- makedata(nobs=1000)
delta <- censoring(y[,1], p.cens)
for (nobs in c(20,100,1000)) {
  yy <- y[1:nobs,2]
  ddelta <- delta[1:nobs]
  ysurv <- Surv(yy, ddelta)
  yfit <- summary(survfit(ysurv~1))
  NAest1 <- cumsum(yfit$n.event/yfit$n.risk) # approach 1
  NAvar1 <- cumsum((yfit$n.risk-yfit$n.event)*yfit$n.event/yfit$n.risk^3) # variance approach 1
  NAest2 <- NAestim2(yfit) # approach 2
  NAvar2 <- NAvaria2(yfit) # variance approach 2
  plot(yfit$time, NAest1-NAest2, type='S', col=4, xlab='time', ylab='difference of estimators', main=paste('n.obs=', nobs, ', p.cens=',p.cens), ylim=1e-2*c(-1,1))
  lines(yfit$time, NAest1 + NAvar2^(1/2) - NAest2 - NAvar2^(1/2), type='S', lty=3, col=4)
  lines(yfit$time, NAest1 - NAvar2^(1/2) - NAest2 + NAvar2^(1/2), type='S', lty=3, col=4)
}
```

Not sure what this is for:
```{r}
par(mfrow=c(1,3))
lambda <- 1
p.cens <- 0.3

y <- makedata(nobs=1e6, lambda=lambda)
delta <- censoring(y[,1], p.cens)
for (nobs in c(1e2,1e4,1e6)) {
  #for (p.cens in c(0,0.2,0.8)) {
    yy <- y[1:nobs,2]
    ddelta <- delta[1:nobs]
    ysurv <- Surv(yy, ddelta)
    yfit <- summary(survfit(ysurv~1))
    NAest1 <- cumsum(yfit$n.event/yfit$n.risk) # approach 1
    NAvar1 <- cumsum((yfit$n.risk-yfit$n.event)*yfit$n.event/yfit$n.risk^3) # variance approach 1
    NAest2 <- NAestim2(yfit) # approach 2
    NAvar2 <- NAvaria2(yfit) # variance approach 2
    plot(yfit$time, NAest1, type='S', col=4, xlab='time', ylab='cumulative hazard', main=paste('n.obs=', nobs, ', p.cens=',p.cens))
    lines(yfit$time, NAest2, type='S', col='purple')
    lines(yfit$time, -log(yfit$surv), type='S', col=2)
    abline(0,lambda, col='grey')
  #}
}
```

This is the good one:
```{r}
par(mfrow=c(3,3))
lambda <- 1
y <- makedata(nobs=1e6, lambda=lambda, digits=5)
length(unique(y[,2]))/length(y[,2])

for (nobs in c(1e2,1e4,1e6)) {
  for (p.cens in c(0,0.2,0.8)) {
    dig = ifelse(nobs==1e2, 1,ifelse(nobs==1e4, 3, 5))
    y <- makedata(nobs=1e6, lambda=lambda, digits=dig) #roughly 30% unique times in each case
    delta <- censoring(y[,1], p.cens)
    yy <- y[1:nobs,2]
    ddelta <- delta[1:nobs]
    ysurv <- Surv(yy, ddelta)
    yfit <- summary(survfit(ysurv~1))
    NAest1 <- cumsum(yfit$n.event/yfit$n.risk) # approach 1
    NAvar1 <- cumsum((yfit$n.risk-yfit$n.event)*yfit$n.event/yfit$n.risk^3) # variance approach 1
    NAest2 <- NAestim2(yfit) # approach 2
    NAvar2 <- NAvaria2(yfit) # variance approach 2
    plot(yfit$time, NAest1, type='S', col=4, xlab='time', ylab='cumulative hazard', main=paste('n.obs=', nobs, ', p.cens=',p.cens))
    lines(yfit$time, NAest2, type='S', col='purple')
    lines(yfit$time, -log(yfit$surv), type='S', col=2)
    abline(0,lambda*(1-p.cens), col='grey')
  };
}
```
Now try with ggplot2:
```{r}
lambda <- 1
y <- makedata(nobs=1e6, lambda=lambda, digits=5)
nobs <- c(1e2,1e4,1e6)
p.cens <- c(0,0.2,0.8)

NOBS <- numeric(0)
PCENS <- numeric(0)
NAest1 <- numeric(0)
NAest2 <- numeric(0)
NAvar1 <- numeric(0)
NAvar2 <- numeric(0)
ytime <- numeric(0)
ySurv <- numeric(0)

for (i in 1:3) {
  for (j in 1:3) {
    delta <- censoring(y[,1], p.cens[j])
    yy <- y[1:nobs[i],2]
    ddelta <- delta[1:nobs[i]]
    ysurv <- Surv(yy, ddelta)
    yfit <- summary(survfit(ysurv~1))
    ySurv <- c(ySurv, yfit$surv)
    ytime <- c(ytime, yfit$time)
    NAest1 <- c(NAest1, cumsum(yfit$n.event/yfit$n.risk)) # approach 1
    NAvar1 <- c(NAvar1, cumsum((yfit$n.risk-yfit$n.event)*yfit$n.event/yfit$n.risk^3)) # variance approach 1
    NAest2 <- c(NAest2, NAestim2(yfit)) # approach 2
    NAvar2 <- c(NAvar2, NAvaria2(yfit)) # variance approach 2
    # plot(yfit$time, NAest1, type='S', col=4, xlab='time', ylab='cumulative hazard', main=paste('n.obs=', nobs, ', p.cens=',p.cens))
    # lines(yfit$time, NAest2, type='S', col='purple')
    # lines(yfit$time, -log(yfit$surv), type='S', col=2)
    # abline(0,lambda*(1-p.cens), col='grey')
    NOBS <- c(NOBS, rep(nobs[i], length(yfit$time)))
    PCENS <- c(PCENS, rep(p.cens[j], length(yfit$time)))
  }
}

df <- data.frame(KM=-log(ySurv), NA1=NAest1, NA2=NAest2, var1=NAvar1, var2=NAvar2, time=ytime, n.obs=NOBS, p.cens=PCENS)

ggplot(df, aes(time)) +
  geom_step(aes(y=KM, color="Kaplan-Meier")) +
  geom_step(aes(y=NA2, color="Nelson-Aalen")) +
  facet_wrap(~n.obs+p.cens, scales="free", labeller=label_both) +
  ylab("estimated cumulative hazard") +
  ggtitle("Kaplan-Meier versus tie-adjusted Nelson-Aalen estimates, varying sample size and amount of censoring")
```


And compare with the no-ties version: yes, we still have the problem with the slopes.
```{r}
par(mfrow=c(3,3))
lambda <- 1
y <- makedata(nobs=1e6, lambda=lambda)

for (nobs in c(1e2,1e4,1e6)) {
  for (p.cens in c(0,0.2,0.8)) {
    delta <- censoring(y[,1], p.cens)
    yy <- y[1:nobs,1]
    ddelta <- delta[1:nobs]
    ysurv <- Surv(yy, ddelta)
    yfit <- summary(survfit(ysurv~1))
    NAest1 <- cumsum(yfit$n.event/yfit$n.risk) # approach 1
    NAvar1 <- cumsum((yfit$n.risk-yfit$n.event)*yfit$n.event/yfit$n.risk^3) # variance approach 1
    NAest2 <- NAestim2(yfit) # approach 2
    NAvar2 <- NAvaria2(yfit) # variance approach 2
    plot(yfit$time, NAest1, type='S', col=4, xlab='time', ylab='cumulative hazard', main=paste('n.obs=', nobs, ', p.cens=',p.cens))
    lines(yfit$time, NAest2, type='S', col='purple')
    lines(yfit$time, -log(yfit$surv), type='S', col=2)
    abline(0,lambda, col='grey')
  }
}
```

## Compare tied and untied versions of N-A
First on data that is really continuous (as in approach 2):
```{r}
set.seed(123)
par(mfrow=c(1,1))
p.cens <- 0.3
y <- makedata(nobs=1e3, lambda=1, digits=2)
delta <- censoring(y[,1], 0.3)

y1 <- y[,1]
y2 <- y[,2]
ysurv1 <- Surv(y1, delta)
ysurv2 <- Surv(y2, delta)
yfit1 <- summary(survfit(ysurv1~1))
yfit2 <- summary(survfit(ysurv2~1))
NAest1 <- cumsum(yfit1$n.event/yfit1$n.risk) #no ties
NAest21 <- cumsum(yfit2$n.event/yfit2$n.risk) #tied approach 1
NAest22 <- NAestim2(yfit2) #tied approach 2
NAvar21 <- cumsum((yfit2$n.risk-yfit2$n.event)*yfit2$n.event/yfit2$n.risk^3) #variance tied 1
NAvar22 <- NAvaria2(yfit2) #variance tied 2

# plot(yfit1$time, NAest1, type='S', col=2, xlab='time', ylab='cumulative hazard', main="Nelson-Aalen and variants for tied and untied data")
# lines(yfit2$time, NAest21, type='S', col=4)
# lines(yfit2$time, NAest22, type='S', col='purple')
# abline(0, 1-p.cens, col='gray', lty=2)
# legend("topleft", c("no ties","approach 1", "approach 2", "true"), col=c(2,4,'purple','grey'), lty=c(1,1,1,2))

plot(yfit1$time, NAest1-yfit1$time*(1-p.cens), type='l', col=3, xlab='time', ylab='estimation error', main="Difference of Nelson-Aalen estimators from true cumulative hazard (rounded data)", ylim=c(-0.4,0.4))
lines(yfit2$time, NAest21-yfit2$time*(1-p.cens), type='l', col=4)
lines(yfit2$time, NAest22-yfit2$time*(1-p.cens), type='l', col='purple')
abline(h=0, col='grey', lty=2)
legend("topleft", c("no ties","approach 1", "approach 2"), col=c(3,4,'purple'), lty=1)
```
Try with ggplot2:
```{r}
library(ggplot2)

df <- data.frame(time=yfit2$time, NA1=NAest21, NA2=NAest22, true=yfit2$time*(1-p.cens), var1=NAvar21, var2=NAvar22)

ggplot(df, aes(x=time)) +
  geom_line(aes(y=NA1-true, color="NA1")) + 
  geom_line(aes(y=NA2-true, color="NA2")) +
  geom_ribbon(aes(ymin=NA1-true-var1^(1/2), ymax=NA1-true+var1^(1/2), fill="NA1"), alpha=0.2, show.legend = FALSE) +
  geom_ribbon(aes(ymin=NA2-true-var2^(1/2), ymax=NA2-true+var2^(1/2), fill="NA2"), alpha=0.2, show.legend = FALSE) +
  ylab("estimation error") + ggtitle("Difference of estimators from true cumulative hazard (rounded continuous data)")

```

Now on data that is really discrete (as in approach 1):
Actually this bit is obsolete, skip to next chunk.
```{r}
pnb <- 0.12
rnb <- 1
p.cens <- 0.3
set.seed(632)

y <- rnbinom(1e2,rnb,pnb)
delta <- censoring(y, p.cens)
ysurv <- Surv(y, delta)
yfit <- summary(survfit(ysurv~1))
NAest21 <- cumsum(yfit$n.event/yfit$n.risk)
NAest22 <- NAestim2(yfit)

plot(yfit$time, NAest21, type='S', col=4, xlab='time', ylab='cumulative hazard', main="Nelson-Aalen estimators (discrete data)")
lines(yfit$time, NAest22, type='S', col='purple')
x <- seq(0, max(y), 1)
trueline <- -pnbinom(x, rnb,pnb, lower.tail=FALSE, log=TRUE) #this is just a straight line
lines(x, trueline*(1-p.cens), col='grey', lty=2)
legend("topleft", c("approach 1", "approach 2","true"), col=c(4,'purple','grey'), lty=c(1,1,2))

# ## not quite right, as true line only subtracted at event times...
# plot(yfit$time, NAest21-yfit$time*(1-p.cens), type='S', col=4, xlab='time', ylab='estimation error', main="Difference of Nelson-Aalen estimators from true cumulative hazard (discrete data)")
# lines(yfit$time, NAest22-yfit$time*(1-p.cens), type='S', col='purple')
# abline(h=0, col='grey', lty=2)
# legend("topleft", c("approach 1", "approach 2"), col=c(4,'purple'), lty=c(1,1))
```

Am I being stupid...? A: yes. Let's use Geometric distn.
```{r}
pnb <- 0.01
rnb <- 1
p.cens <- 0.3
set.seed(321)

y <- rgeom(1e3,pnb)
delta <- censoring(y, p.cens)
ysurv <- Surv(y, delta)
yfit <- summary(survfit(ysurv~1))
NAest21 <- cumsum(yfit$n.event/yfit$n.risk) #approach 1
NAest22 <- NAestim2(yfit) #approach 2
NAvar21 <- cumsum((yfit$n.risk-yfit$n.event)*yfit$n.event/yfit$n.risk^3) # variance approach 1
NAvar22 <- NAvaria2(yfit) # variance approach 2

# plot(yfit$time, NAest21, type='S', col=4, xlab='time', ylab='cumulative hazard', main="Nelson-Aalen estimators (discrete data)")
# lines(yfit$time, NAest22, type='S', col='purple')
# x <- 0:max(y)
# #trueline <- -pgeom(x, pnb, lower.tail=FALSE, log=TRUE) #this is just a straight line
# #lines(x, trueline*(1-p.cens), col='grey', lty=2)
# lines(x, -x*log(1-pnb)*(1-p.cens), col='grey', lty=2)
# legend("topleft", c("approach 1", "approach 2","true"), col=c(4,'purple','grey'), lty=c(1,1,2))
# # also add CIs:
# lines(yfit$time, NAest21+NAvar21, type='S', col=4, lty=3)
# lines(yfit$time, NAest21-NAvar21, type='S', col=4, lty=3)
# lines(yfit$time, NAest22+NAvar22, type='S', col='purple', lty=3)
# lines(yfit$time, NAest22-NAvar22, type='S', col='purple', lty=3)

plot(yfit$time, NAest21+yfit$time*log(1-pnb)*(1-p.cens), type='l', col=4, xlab='time', ylab='estimation error', main="Difference of Nelson-Aalen estimators from true cumulative hazard (discrete data)", ylim=c(-0.2,0.65))
lines(yfit$time, NAest22+yfit$time*log(1-pnb)*(1-p.cens), type='l', col='purple')
abline(h=0, col='grey', lty=2)
legend("topleft", c("approach 1", "approach 2"), col=c(4,'purple'), lty=c(1,1))
# to add also CIs:
lines(yfit$time, NAest21+NAvar21+yfit$time*log(1-pnb)*(1-p.cens), type='l', col=4, lty=3)
lines(yfit$time, NAest21-NAvar21+yfit$time*log(1-pnb)*(1-p.cens), type='l', col=4, lty=3)
lines(yfit$time, NAest22+NAvar22+yfit$time*log(1-pnb)*(1-p.cens), type='l', col='purple', lty=3)
lines(yfit$time, NAest22-NAvar22+yfit$time*log(1-pnb)*(1-p.cens), type='l', col='purple', lty=3)
```

Try with ggplot2:
```{r}
library(ggplot2)

df <- data.frame(time=yfit$time, NA1=NAest21, NA2=NAest22, true=-yfit$time*log(1-pnb)*(1-p.cens), var1=NAvar21, var2=NAvar22)

ggplot(df, aes(x=time)) +
  geom_line(aes(y=NA1-true, color="NA1")) + 
  geom_line(aes(y=NA2-true, color="NA2")) +
  geom_ribbon(aes(ymin=NA1-true-var1^(1/2), ymax=NA1-true+var1^(1/2), fill="NA1"), alpha=0.2, show.legend = FALSE) +
  geom_ribbon(aes(ymin=NA2-true-var2^(1/2), ymax=NA2-true+var2^(1/2), fill="NA2"), alpha=0.2, show.legend = FALSE) +
  ylab("estimation error") + ggtitle("Difference of estimators from true cumulative hazard (discrete data)")

```

Example for presentation:
```{r}
T <- c(0, 0.9, 3.8, 4.1,5)
Tprime <- round(T, digits=0)
N.A <- c(0, 0.2, 0.45, 0.738, 0.738)
Disc <- c(0, 0.2, 0.7, 0.7, 0.7)
Cts <- c(0, 0.2, 0.738, 0.738, 0.738)


par(mfrow=c(2,1),fig=c(0.1,0.9,0.8,0.9), mai=c(0.1,0.1,0.1,0.1))
plot(NA,xlim=c(0,5), ylim=c(0,1),axes=F,ann=F);
axis(1)
points(T[2:4],rep(0.45,length(T[2:4])), col=1, pch=16)
text(T[2:4], 0.2, c("T1", "T2", "T3"))
points(Tprime[2:4],rep(0.65,length(Tprime[2:4])), col=2, pch=16)
text(Tprime[2:3], 0.9, c("T1", "T2=T3"), col=2)

par(fig=c(0.1,0.9,0.1,0.7), new=TRUE)
plot(T, N.A, type="s", col=1, lwd=2, xlab="time", ylab="cumulative hazard etsimate")
legend("bottomright", c("N-A", "discrete", "contiuous"), col=c(1,2,4), lwd=2)
points(Tprime, Disc, type="s", col=2, lwd=2)
points(Tprime, Cts, type="s", col=4, lwd=2)
```